Depression and suicidality as costly signals of need

### 2021-03-22 Paper.rmd
Added to the discussion

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the liklihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. However, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness participants had with the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family or a romantic partner). With the thwarted marriage vignette, the fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality (see limitations for more information).

In the "brother-in-law" and "romantic partner" scenarios, there was a noticeable jump in the magnitude of pro-signaler responses when going from the crying to mild depression and then a second, smaller jump from depression to a sucide attempt (Ed please check this is the right interpretation. It's definitely not the most conservative one). One potential reason for this is the increase in symptoms from mild depression to depression was not costly enough to result in a large effect as the symptoms only differed in severity (e.g., grades dropping from As to Bs [mild depression] or Cs [depression]).

@@

Our prediction that the effect of signal cost on likelihood of help would largely be mediated by signal's effect on belief in need was also met, as would be expected if the main function of signaling need is a reduction in the amount of private information one has relative to those motivated to care for the signaler (cite). However, there was still a positive main effect of the signal in many pairwise comparisons between the costlier signals and the less costly signals, including the comparison between the depression and verbal request condition (See SI for other mediation models). One possible reason for this is that someone may still be distressed and face real adversity without being forthcoming about one's problems. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, with stronger signals still conveying the severity of the competition between her and her cousin in ways words alone could not.

@@

Scenarios likely favor cheating due to high potential of benefits

>>>>

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the likelihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. However, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness between participants and the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family or a romantic partner). With the thwarted marriage vignette, the fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality, limiting our confidence in this interpretation (see limitations for more information).

In the "brother-in-law" and "romantic partner" scenarios, there was a noticeable jump in the magnitude of pro-signaler responses when going from the crying to mild depression and then a second, smaller jump from depression to a suicide attempt. One potential reason for this is the increase in symptoms from mild depression to depression was not costly enough to result in a large effect as the symptoms only differed in severity (e.g., grades dropping from As to Bs [mild depression] or Cs [depression]). Such small changes may be less impactful in vignette form than in real-life scenarios in which the effects of signaling may increase in severity and be prolonged over a lenghty period of time.

@@

Our prediction that the effect of signal cost on likelihood of help would largely be mediated by signal's effect on belief in need was also met, as would be expected if the main function of signaling need is a reduction in the amount of private information one has relative to those motivated to care for the signaler (cite). However, there was still a positive main effect of the signal in many pairwise comparisons between the costlier signals and the less costly signals, including the comparison between the depression and verbal request condition (See SI for other mediation models). One possible reason for this is that someone may still be distressed and face real adversity without being forthcoming about one's problems. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, but the stronger signals may have still conveyed the severity of the competition between her and her cousin in ways words alone could not.

@@

Our results on the connection between belief in need and likelihood of help are consistent with the bargaining models of depression and suicide. In addition to the effect of signal cost on support being largely mediated by the signal's effect on belief in need, belief in need and likelihood of support were highly correlated but distinct at T1 (pre-signal). This situation should arise in any instance when a target of a signal of need would benefit from helping if the need was genuine and known to reasonable precision. Importantly, such problems are common among humans today (cite) and have likely been so since individuals could increase their inclusive fitness through supporting kin or important cooperative partners in times of need or conflict over acceptable one's contributions joint endeavors (cite).

Although this study was not designed to examine whether the costs of signaling would be made up for by the increased support stemming from depression and suicide attempts directly, the results are encouraging. In general, the costs required to honestly convey one's qualities should be expected to increase with the potential payoffs of deception (cite), with the vignettes in this study emphasizing the existence of private information and sizable conflicts of interest to increase the perceived risk of being taken advantage of. In similar real-world scenarios, this difficulty of signaling honest need would coincide with the substantial fitness-relevant benefits of preventing future assaults or being able to marry the partner of one's choice. These benefits coupled with the fact (too strong?) that depression and suicide attempts become relatively less costly with increased need (Hagen 2003), suggest that the increased support seen in this study may be worth the costs required to signal need more often than is often appreciated when counting the entire costs of depression as the signal rather than only the upregulation of symptoms for a communicative purpose.
